Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-12 latex/main.tex
Incoroporate the tanglegrams

The ARG represents the full evolutionary history of SARS-CoV-2 and 
is in close agreement with a state-of-the-art UShER phylogeny built using the Viridian dataset \cite{Hunt2024}.
First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024}
by pruning both down to representative samples for 1097 deeply-sampled lineages (STAR Methods).
Tanglegrams reveal that the ARG captures known phylogenetic relationships
both within sublineages and at the level of major lineages (including VOCs),
with good agreement between the phylogenetic bipartitions
in the ARG and the UShER phylogeny (Figure~\ref{fig:phylo_backbones}).
Next, we compared the sc2ts and UShER inferences 
from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites,
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG against 
the mutational spectra 
from the Viridian UShER phylogeny \cite{Hunt2024} and a previous study \cite{Bloom2023} (STAR Methods),
finding a close match to them (Figure~\ref{fig:mutational_spectra}).

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
% Referenced by the main Results.
The ARG captures highly prevalent deletions in SARS-CoV-2 (Table~\ref{tab:major_dels}).

<TODO> Highly recurrent deletions (Table~\ref{tab:rec_dels}).

@@

\begin{table}
\centering
\caption{Major deletion events in the ARG.}
\label{tab:major_dels}
\begin{tabular}{rrcrcrr}
\toprule
Start & Length & Region & Node & Pango lineage & Count & Frequency (\%)\\
\midrule
28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 \\
28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 \\
22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112504 & 44.8200 \\
11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 \\
28362 & 9 & N & 1436802 & B.1.1.529 & 871774 & 35.1216 \\
21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 532072 & 21.4359 \\
11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 531894 & 21.4287 \\
6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 340757 & 13.7283 \\
11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 339867 & 13.6924 \\
21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339681 & 13.6849 \\
21987 & 9 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339651 & 13.6837 \\
22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 312706 & 12.5982 \\
28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 290446 & 11.7014 \\
11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 290408 & 11.6998 \\
21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 288831 & 11.6363 \\
21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 290438 & 11.7010 \\
21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 190425 & 7.6718 \\
686 & 9 & ORF1ab / NSP1 & 2698748 & BA.4 (Omicron) & 23330 & 0.9399 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{Highly recurrent deletions.}
\label{tab:rec_dels}
\begin{tabular}{rrcr}
\toprule
Start & Length & Region & Occurrences \\
\midrule
29781 & 1 & non-coding & 12754 \\
29762 & 1 & non-coding & 11383 \\
29779 & 1 & non-coding & 10951 \\
29769 & 1 & non-coding & 10809 \\
29774 & 1 & non-coding & 10735 \\
29700 & 1 & non-coding & 8562 \\
686 & 9 & ORF1ab / NSP1 & 6436 \\
29614 & 1 & ORF10 & 5039 \\
29555 & 1 & non-coding & 4814 \\
29543 & 1 & non-coding & 4804 \\
29541 & 1 & non-coding & 4802 \\
203 & 1 & non-coding & 3604 \\
222 & 1 & non-coding & 3442 \\
21991 & 3 & S / S1-NTD & 2088 \\
28254 & 1 & ORF8 & 1969 \\
28093 & 1 & ORF8 & 900 \\
21765 & 6 & S / S1-NTD & 482 \\
27532 & 1 & ORF7a & 404 \\
335 & 1 & ORF1ab / NSP1 & 370 \\
337 & 1 & ORF1ab / NSP1 & 346 \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
\centering
% \includegraphics{sc2ts_methodology}
\caption{
Phylogenetic backbones: TODO
}
\label{fig:phylo_backbones}
\end{figure}

>>>>

The ARG represents the full evolutionary history of SARS-CoV-2 and 
is in close agreement with a state-of-the-art UShER phylogeny built using the Viridian dataset \cite{Hunt2024}.
First, we validated the broad structure of the ARG against the UShER phylogeny \cite{Hunt2024}
by pruning both down to representative samples for 1097 deeply-sampled 
lineages (STAR Methods).
Tanglegrams show that the ARG closely matches the phylogenetic
structure of the UShER tree (Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
Next, we compared the sc2ts and UShER inferences 
from the Viridian dataset
by pruning both to their intersection of 2,475,418 samples and 27,507 sites,
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
and found they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG against 
the mutational spectra 
from the Viridian UShER phylogeny \cite{Hunt2024} and a previous study \cite{Bloom2023} (STAR Methods),
finding a close match to them (Figure~\ref{fig:mutational_spectra}).

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
% Referenced by the main Results.
The ARG captures important deletion events (Table~\ref{tab:major_dels}) and highly recurrent deletion events (Table~\ref{tab:rec_dels}).

@@

\begin{table}
\centering
\caption{Major deletion events in the ARG. These events are inherited by at least 10,000 samples in the ARG. Count is the number of samples inheriting a deletion. The Pango lineage assigned to the node above which an event occurs is shown. Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in Subunit 1.}
\label{tab:major_dels}
\begin{tabular}{rrcrcrr}
\toprule
Start & Length & Region & Node & Pango lineage & Count & Frequency (\%)\\
\midrule
28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 \\
28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 \\
22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112504 & 44.8200 \\
11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 \\
28362 & 9 & N & 1436802 & B.1.1.529 & 871774 & 35.1216 \\
21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 532072 & 21.4359 \\
11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 531894 & 21.4287 \\
6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 340757 & 13.7283 \\
11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 339867 & 13.6924 \\
21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339681 & 13.6849 \\
21987 & 9 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339651 & 13.6837 \\
22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 312706 & 12.5982 \\
28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 290446 & 11.7014 \\
11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 290408 & 11.6998 \\
21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 288831 & 11.6363 \\
21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 290438 & 11.7010 \\
21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 190425 & 7.6718 \\
686 & 9 & ORF1ab / NSP1 & 2698748 & BA.4 (Omicron) & 23330 & 0.9399 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{Highly recurrent deletions. These are the top 20 deletions in terms of the number of occurrences in the ARG. Abbreviations: NSP, non-structural protein; ORF, open reading frame; S, Spike; S1-NTD, N-terminal domain in Subunit 1.}
\label{tab:rec_dels}
\begin{tabular}{rrcr}
\toprule
Start & Length & Region & Occurrences \\
\midrule
29781 & 1 & non-coding & 12754 \\
29762 & 1 & non-coding & 11383 \\
29779 & 1 & non-coding & 10951 \\
29769 & 1 & non-coding & 10809 \\
29774 & 1 & non-coding & 10735 \\
29700 & 1 & non-coding & 8562 \\
686 & 9 & ORF1ab / NSP1 & 6436 \\
29614 & 1 & ORF10 & 5039 \\
29555 & 1 & non-coding & 4814 \\
29543 & 1 & non-coding & 4804 \\
29541 & 1 & non-coding & 4802 \\
203 & 1 & non-coding & 3604 \\
222 & 1 & non-coding & 3442 \\
21991 & 3 & S / S1-NTD & 2088 \\
28254 & 1 & ORF8 & 1969 \\
28093 & 1 & ORF8 & 900 \\
21765 & 6 & S / S1-NTD & 482 \\
27532 & 1 & ORF7a & 404 \\
335 & 1 & ORF1ab / NSP1 & 370 \\
337 & 1 & ORF1ab / NSP1 & 346 \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}
\centering
\centering
\includegraphics[width=\linewidth]{tanglegram_base_tree.pdf}
\caption{
Tanglegram comparing phylogenetic backbones of sc2ts and UShER.
}
\label{fig:tanglegram_base}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{tanglegram_Delta_subtree.pdf}
\caption{
Tanglegram comparing sc2ts and UShER on the Delta subtree.
}
\label{fig:tanglegram_delta}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{tanglegram_BA.2_subtree.pdf}
\caption{
Tanglegram comparing sc2ts and UShER on the BA.2 subtree.
}
\label{fig:tanglegram_ba2}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{tanglegram_BA.5_subtree.pdf}
\caption{
Tanglegram comparing sc2ts and UShER on the BA.5 subtree.
}
\label{fig:tanglegram_ba5}
\end{figure}

\begin{figure}
\centering
\centering
\includegraphics[width=\linewidth]{subgraph-Alpha.pdf}
\caption{
Subgraph of Alpha
}
\label{fig:subgraph_alpha}
\end{figure}

\begin{figure}
\centering
\centering
\includegraphics[width=\linewidth]{subgraph-Delta.pdf}
\caption{
Subgraph of Delta
}
\label{fig:subgraph_delta}
\end{figure}

\begin{figure}
\centering
\centering
\includegraphics[width=\linewidth]{subgraph-BA.1+BA.2.pdf}
\caption{
Subgraph of BA.1 and BA.2
}
\label{fig:subgraph_ba1_ba2}
\end{figure}
